---
title: "ggmultivar: Multivariate Analysis Plots Using ggplot2"
output:
  md_document:
    variant: gfm
    preserve_yaml: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
```

[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT) [![R-CMD-check](https://github.com/JannikVindeloev/ggmultivar/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JannikVindeloev/ggmultivar/actions/workflows/R-CMD-check.yaml)

**ggmultivar** provides a tidyverse-style interface for creating scores and loading plots for multivariate analysis methods including PCA, PLS, sPLS, and sPLS-DA. The package is designed to work seamlessly with both native R data structures and [mixOmics](https://mixomicsteam.github.io/) objects.

## Features

- **Tidyverse-compatible**: All functions follow tidyverse conventions and return tidy data frames
- **ggplot2-based**: Create publication-quality plots using the familiar ggplot2 syntax
- **mixOmics compatibility**: Directly handle mixOmics objects (PCA, PLS, sPLS, sPLS-DA, DIABLO, RGCCA)
- **Comprehensive plotting**: Scores plots, loadings plots, and biplots for all supported methods
- **Flexible customization**: Color, shape, and size points by external variables
- **Statistical annotations**: Automatic variance explained labels, confidence ellipses

## Installation

```r
# Install from GitHub
devtools::install_github("JannikVindeloev/ggmultivar")

# Or install dependencies first
install.packages(c("ggplot2", "dplyr", "tidyr", "purrr", "tibble"))
devtools::install_github("JannikVindeloev/ggmultivar")
```

## Quick Start

### Basic Usage with Base R Data

```{r}
library(ggmultivar)

# Create sample scores data
scores_data <- data.frame(
  sample = paste0("Sample_", 1:50),
  component = rep(1:2, each = 50),
  score = c(rnorm(50), rnorm(50))
)

# Create sample loadings data
loadings_data <- data.frame(
  variable = paste0("Var_", 1:10),
  component = rep(1:2, each = 10),
  loading = rnorm(20)
)

# Create scores plot
p1 <- ggscores(scores_data) +
  ggtitle("Basic Scores Plot")

# Create loadings plot
p2 <- ggloadings(loadings_data) +
  ggtitle("Basic Loadings Plot")

# Create biplot
p3 <- ggbiplot(list(scores = scores_data, loadings = loadings_data)) +
  ggtitle("Basic Biplot")
```

### With Coloring and Grouping

```{r}
# Add grouping variable
color_vec <- sample(letters[1:3], 50, replace = TRUE)

# Scores plot with coloring
p4 <- ggscores(scores_data, color_by = color_vec) +
  geom_encircle(group_by = "color_by") +
  ggtitle("Scores Plot with Groups")

# Loadings plot with colored arrows
p5 <- ggloadings(loadings_data, color_by = abs(loadings_data$loading)) +
  ggtitle("Loadings Plot with Colored Arrows")

# Biplot with grouping
p6 <- ggbiplot(
  list(scores = scores_data, loadings = loadings_data),
  color_by = color_vec
) +
  geom_encircle(group_by = "color_by") +
  ggtitle("Biplot with Groups")
```

### With Real Data (Iris)

```{r}
# Using iris dataset
data(iris)

# Create a simple PCA-like visualization
# Note: For actual PCA, use prcomp() or mixOmics::pca()
scores_iris <- data.frame(
  sample = rownames(iris),
  component = rep(1:2, each = nrow(iris)),
  score = c(as.numeric(iris$Sepal.Length), as.numeric(iris$Sepal.Width))
)

p7 <- ggscores(scores_iris, color_by = iris$Species) +
  geom_encircle(group_by = "color_by") +
  ggtitle("Iris Data - Scores Plot by Species")
```

### With mixOmics Objects

If you have mixOmics installed, you can use it with datasets from FactoMineR:

```r
# Install required packages
# install.packages(c("mixOmics", "FactoMineR"))

library(mixOmics)
library(FactoMineR)
data(wine)

# PCA with mixOmics
pca_mix <- pca(wine$X, ncomp = 3)

# Create scores plot
ggscores(pca_mix, color_by = wine$Y) +
  geom_encircle(group_by = "color_by") +
  ggtitle("PCA Scores Plot - Wine Data")

# Create loadings plot
ggloadings(pca_mix) +
  geom_correlation_circle() +
  ggtitle("PCA Loadings Plot")

# Create biplot
ggbiplot(pca_mix, color_by = wine$Y) +
  geom_encircle(group_by = "color_by") +
  ggtitle("PCA Biplot")

# PLS with mixOmics
pls_mix <- pls(wine$X, wine$Y, ncomp = 3)
ggbiplot(pls_mix)

# sPLS-DA with mixOmics
splsda_mix <- splsda(wine$X, wine$Y, ncomp = 3)
ggscores(splsda_mix)
```

## Function Overview

### Core Plotting Functions

- `ggscores()`: Create scores plot from data frame or mixOmics object
- `ggloadings()`: Create loadings plot from data frame or mixOmics object
- `ggbiplot()`: Create combined scores and loadings plot

### Custom Geoms

- `geom_encircle()`: Add ellipses or convex hulls to encircle groups
- `geom_correlation_circle()`: Add correlation circles to biplots
- `geom_confidence_interval()`: Add confidence intervals to loading vectors
- `geom_sample_labels()`: Add sample labels to scores plots
- `geom_variable_labels()`: Add variable labels to loadings plots

### S3 Methods for mixOmics Objects

- `scores()`: Extract scores from mixOmics objects
- `loadings()`: Extract loadings from mixOmics objects
- `explained_variance()`: Extract explained variance from mixOmics objects

## Plot Customization

All plotting functions accept standard ggplot2 arguments and provide additional customization options:

```r
# Customize scores plot
ggscores(scores_data, 
         x_component = 1, 
         y_component = 2,
         color_by = color_vec,
         shape_by = sample(letters[1:2], 50, replace = TRUE),
         show_ellipse = TRUE,
         ellipse_level = 0.95,
         show_labels = TRUE,
         label_size = 3,
         theme = theme_bw()) +
  scale_color_brewer(palette = "Set1") +
  ggtitle("Customized Scores Plot")

# Customize loadings plot
ggloadings(loadings_data,
           show_arrows = TRUE,
           arrow_length = 0.8,
           show_circle = TRUE,
           circle_radius = 1,
           color_by = abs(loadings_data$loading))
```

## Output Format

All analysis functions return a list with the following components:

- `scores`: Tidy data frame with component scores for each sample
- `loadings`: Tidy data frame with component loadings for each variable
- `explained_variance`: Data frame with variance explained by each component
- Additional method-specific components (e.g., `y_loadings`, `weights`, `r2`)

## Supported Methods

- **PCA**: Principal Component Analysis
- **PLS**: Partial Least Squares Regression
- **sPLS**: Sparse Partial Least Squares
- **sPLS-DA**: Sparse Partial Least Squares Discriminant Analysis
- **DIABLO**: Multi-omics integration (via mixOmics)
- **RGCCA**: Regularized Generalized Canonical Correlation Analysis (via mixOmics)

## Dependencies

- Required: R (>= 3.6.0), ggplot2, dplyr, tidyr, purrr, tibble
- Suggested: mixOmics, FactoMineR, testthat, knitr, rmarkdown, covr

## Contributing

Contributions are welcome! Please feel free to submit issues or pull requests.

## License

MIT License
