---
title: "Introduction to ggmultivar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ggmultivar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

# Introduction to ggmultivar

`ggmultivar` is an R package that provides a tidyverse-style interface for creating scores and loading plots for multivariate analysis methods. This vignette demonstrates the main functionality of the package.

## Installation

```r
# Install from GitHub
devtools::install_github("JannikVindeloev/ggmultivar")
```

## Loading the package

```{r}
library(ggmultivar)
```

## Principal Component Analysis (PCA)

### Basic PCA with mtcars data

Let's start with a simple PCA example using the built-in `mtcars` dataset.

```{r}
# Perform PCA on numeric columns
data(mtcars)
pca_result <- perform_pca(mtcars[, 1:7], n_components = 3)

# View the structure of the result
str(pca_result)

# View explained variance
pca_result$explained_variance
```

### Creating Scores Plots

```{r}
# Basic scores plot
p1 <- pca_scores_plot(pca_result)
print(p1)

# Scores plot with coloring by cylinder
p2 <- pca_scores_plot(pca_result, color_by = mtcars$cyl) +
  ggtitle("PCA Scores Plot - Colored by Cylinder")
print(p2)

# Scores plot with ellipses
p3 <- pca_scores_plot(pca_result, color_by = mtcars$cyl, 
                     show_ellipse = TRUE, ellipse_level = 0.95) +
  ggtitle("PCA Scores Plot with Confidence Ellipses")
print(p3)
```

### Creating Loadings Plots

```{r}
# Basic loadings plot
p4 <- pca_loadings_plot(pca_result)
print(p4)

# Loadings plot with colored arrows
p5 <- pca_loadings_plot(pca_result, 
                        color_by = abs(pca_result$loadings$loading))
print(p5)
```

### Creating Biplots

```{r}
# Basic biplot
p6 <- pca_biplot(pca_result)
print(p6)

# Biplot with coloring by cylinder
p7 <- pca_biplot(pca_result, color_by = mtcars$cyl) +
  ggtitle("PCA Biplot - Colored by Cylinder")
print(p7)
```

## Partial Least Squares (PLS)

### Basic PLS with simulated data

```{r}
# Simulate data
set.seed(123)
X <- matrix(rnorm(50 * 10), nrow = 50, ncol = 10)
Y <- matrix(rnorm(50 * 2), nrow = 50, ncol = 2)

# Perform PLS
pls_result <- perform_pls(X, Y, n_components = 3)

# View the structure of the result
str(pls_result)

# View explained variance
pls_result$explained_variance
```

### Creating PLS Plots

```{r}
# PLS scores plot
p8 <- pls_scores_plot(pls_result)
print(p8)

# PLS loadings plot
p9 <- pls_loadings_plot(pls_result)
print(p9)

# PLS biplot
p10 <- pls_biplot(pls_result)
print(p10)
```

## Working with mixOmics Objects

`ggmultivar` is designed to work seamlessly with [mixOmics](https://mixomicsteam.github.io/) objects. If you have mixOmics installed, you can use the following functions:

```{r, eval = FALSE}
library(mixOmics)
data(wine)

# PCA with mixOmics
pca_mix <- pca(wine$X, ncomp = 3)
ggscores(pca_mix, color_by = wine$Y)

# PLS with mixOmics
pls_mix <- pls(wine$X, wine$Y, ncomp = 3)
ggbiplot(pls_mix)

# sPLS-DA with mixOmics
splsda_mix <- splsda(wine$X, wine$Y, ncomp = 3)
ggscores(splsda_mix)
```

## Customization Options

All plotting functions accept standard ggplot2 arguments and provide additional customization options:

### Common Parameters

- `x_component`, `y_component`: Which components to plot on x and y axes
- `color_by`, `shape_by`, `size_by`: Aesthetic mappings for points
- `show_ellipse`: Whether to show confidence ellipses
- `ellipse_level`: Confidence level for ellipses (0-1)
- `show_labels`: Whether to show sample/variable labels
- `label_size`: Size of labels
- `show_circle`: Whether to show unit circle (for loadings plots)
- `circle_radius`: Radius of unit circle
- `arrow_length`: Scaling factor for loading arrows
- `theme`: ggplot2 theme to apply

### Example: Customized Plot

```{r}
# Customized PCA scores plot
p_custom <- pca_scores_plot(pca_result, 
                          x_component = 1, 
                          y_component = 2,
                          color_by = mtcars$cyl,
                          shape_by = mtcars$gear,
                          show_ellipse = TRUE,
                          ellipse_level = 0.95,
                          show_labels = TRUE,
                          label_size = 3,
                          theme = ggplot2::theme_bw()) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::ggtitle("Customized PCA Scores Plot")

print(p_custom)
```

## Conclusion

This vignette has demonstrated the main functionality of `ggmultivar` for creating multivariate analysis plots using ggplot2 syntax. The package provides:

- Easy-to-use functions for PCA and PLS analysis
- Tidyverse-compatible data structures
- Seamless integration with mixOmics objects
- Flexible customization options
- Publication-quality plots

For more information, see the package documentation and the [GitHub repository](https://github.com/JannikVindeloev/ggmultivar).
