---
title: "Introduction to ggmultivar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ggmultivar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

# Introduction to ggmultivar

`ggmultivar` is an R package that provides a tidyverse-style interface for creating scores and loading plots for multivariate analysis methods. This vignette demonstrates the main functionality of the package using base R datasets.

## Installation

```r
# Install from GitHub
devtools::install_github("JannikVindeloev/ggmultivar")
```

## Loading the package

```{r}
library(ggmultivar)
```

## Getting Started with Base R Data

### Creating Sample Data

For demonstration purposes, we'll create sample scores and loadings data that mimics the output from multivariate analysis methods like PCA or PLS.

```{r}
# Create sample scores data (mimics PCA scores)
set.seed(123)
scores_data <- data.frame(
  sample = paste0("Sample_", 1:50),
  component = rep(1:2, each = 50),
  score = c(rnorm(50, mean = 0, sd = 2), rnorm(50, mean = 0, sd = 2))
)

# Create sample loadings data (mimics PCA loadings)
loadings_data <- data.frame(
  variable = paste0("Var_", 1:10),
  component = rep(1:2, each = 10),
  loading = c(rnorm(10, mean = 0.5, sd = 0.3), rnorm(10, mean = 0.5, sd = 0.3))
)

# View the structure
str(scores_data)
str(loadings_data)
```

### Creating Scores Plots

```{r}
# Basic scores plot
p1 <- ggscores(scores_data)
print(p1)

# Scores plot with coloring by a grouping variable
color_vec <- sample(letters[1:3], 50, replace = TRUE)
p2 <- ggscores(scores_data, color_by = color_vec) +
  ggtitle("Scores Plot - Colored by Group")
print(p2)

# Scores plot with ellipses
p3 <- ggscores(scores_data, color_by = color_vec, 
               show_ellipse = TRUE, ellipse_level = 0.95) +
  ggtitle("Scores Plot with Confidence Ellipses")
print(p3)

# Scores plot with labels
p4 <- ggscores(scores_data, show_labels = TRUE, label_size = 2.5) +
  ggtitle("Scores Plot with Sample Labels")
print(p4)
```

### Creating Loadings Plots

```{r}
# Basic loadings plot
p5 <- ggloadings(loadings_data)
print(p5)

# Loadings plot with colored arrows
p6 <- ggloadings(loadings_data, 
                 color_by = abs(loadings_data$loading)) +
  ggtitle("Loadings Plot - Colored by Magnitude")
print(p6)

# Loadings plot without arrows
p7 <- ggloadings(loadings_data, show_arrows = FALSE, show_labels = TRUE) +
  ggtitle("Loadings Plot - Labels Only")
print(p7)

# Loadings plot without circle
p8 <- ggloadings(loadings_data, show_circle = FALSE) +
  ggtitle("Loadings Plot - No Unit Circle")
print(p8)
```

### Creating Biplots

```{r}
# Basic biplot
p9 <- ggbiplot(list(scores = scores_data, loadings = loadings_data))
print(p9)

# Biplot with coloring by group
p10 <- ggbiplot(list(scores = scores_data, loadings = loadings_data),
                color_by = color_vec) +
  geom_encircle(group_by = "color_by") +
  ggtitle("Biplot - Colored by Group with Ellipses")
print(p10)

# Biplot without loading labels
p11 <- ggbiplot(list(scores = scores_data, loadings = loadings_data),
                show_loading_labels = FALSE) +
  ggtitle("Biplot - No Loading Labels")
print(p11)

# Biplot without circle
p12 <- ggbiplot(list(scores = scores_data, loadings = loadings_data),
                show_circle = FALSE) +
  ggtitle("Biplot - No Unit Circle")
print(p12)
```

## Using Real Data: Iris Dataset

Let's use the built-in `iris` dataset to create more realistic visualizations.

```{r}
# Load iris dataset
data(iris)

# Create scores data from iris (using first two principal components conceptually)
# In practice, you would use prcomp() or mixOmics::pca() for actual PCA
scores_iris <- data.frame(
  sample = rownames(iris),
  component = rep(1:2, each = nrow(iris)),
  score = c(scale(iris$Sepal.Length), scale(iris$Sepal.Width))
)

# Create loadings data (conceptual)
loadings_iris <- data.frame(
  variable = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"),
  component = rep(1:2, each = 2),
  loading = c(0.7, 0.6, -0.5, -0.4)  # Conceptual loadings
)

# Scores plot with iris data
p13 <- ggscores(scores_iris, color_by = iris$Species) +
  geom_encircle(group_by = "color_by") +
  ggtitle("Iris Data - Scores Plot by Species")
print(p13)

# Loadings plot
p14 <- ggloadings(loadings_iris) +
  ggtitle("Iris Data - Loadings Plot")
print(p14)

# Biplot with iris data
p15 <- ggbiplot(list(scores = scores_iris, loadings = loadings_iris),
                color_by = iris$Species) +
  geom_encircle(group_by = "color_by") +
  ggtitle("Iris Data - Biplot by Species")
print(p15)
```

## Custom Geoms

### Encircle Groups

```{r}
# Using the iris scores data
p16 <- ggscores(scores_iris, color_by = iris$Species) +
  geom_encircle(group_by = "color_by", color = "blue", linetype = "solid") +
  ggtitle("Scores Plot with Custom Ellipses")
print(p16)

# Using convex hull
p17 <- ggscores(scores_iris, color_by = iris$Species) +
  geom_encircle(group_by = "color_by", geom = "hull", fill = "red", alpha = 0.1) +
  ggtitle("Scores Plot with Convex Hulls")
print(p17)
```

### Correlation Circle

```{r}
# Loadings plot with correlation circle
p18 <- ggloadings(loadings_data) +
  geom_correlation_circle(radius = 1, color = "blue", linetype = "solid") +
  ggtitle("Loadings Plot with Custom Correlation Circle")
print(p18)
```

## Working with mixOmics Objects

If you have mixOmics installed, you can use it with datasets from FactoMineR:

```{r, eval = FALSE}
# Install required packages if needed
# install.packages(c("mixOmics", "FactoMineR"))

library(mixOmics)
library(FactoMineR)
data(wine)

# PCA with mixOmics
pca_mix <- pca(wine$X, ncomp = 3)

# Create scores plot
ggscores(pca_mix, color_by = wine$Y) +
  geom_encircle(group_by = "color_by") +
  ggtitle("PCA Scores Plot - Wine Data")

# Create loadings plot
ggloadings(pca_mix) +
  geom_correlation_circle() +
  ggtitle("PCA Loadings Plot")

# Create biplot
ggbiplot(pca_mix, color_by = wine$Y) +
  geom_encircle(group_by = "color_by") +
  ggtitle("PCA Biplot")

# PLS with mixOmics
pls_mix <- pls(wine$X, wine$Y, ncomp = 3)
ggbiplot(pls_mix)

# sPLS-DA with mixOmics
splsda_mix <- splsda(wine$X, wine$Y, ncomp = 3)
ggscores(splsda_mix)
```

## Customization Options

All plotting functions accept standard ggplot2 arguments and provide additional customization options:

### Common Parameters

- `x_component`, `y_component`: Which components to plot on x and y axes
- `color_by`, `shape_by`, `size_by`: Aesthetic mappings for points
- `show_ellipse`: Whether to show confidence ellipses
- `ellipse_level`: Confidence level for ellipses (0-1)
- `show_labels`: Whether to show sample/variable labels
- `label_size`: Size of labels
- `show_circle`: Whether to show unit circle (for loadings plots)
- `circle_radius`: Radius of unit circle
- `arrow_length`: Scaling factor for loading arrows
- `theme`: ggplot2 theme to apply

### Example: Customized Plot

```{r}
# Customized scores plot
p_custom <- ggscores(scores_data, 
                    x_component = 1, 
                    y_component = 2,
                    color_by = color_vec,
                    shape_by = sample(letters[1:2], 50, replace = TRUE),
                    show_ellipse = TRUE,
                    ellipse_level = 0.95,
                    show_labels = TRUE,
                    label_size = 2.5,
                    theme = ggplot2::theme_bw()) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::ggtitle("Customized Scores Plot")

print(p_custom)
```

## Conclusion

This vignette has demonstrated the main functionality of `ggmultivar` for creating multivariate analysis plots using ggplot2 syntax with base R datasets. The package provides:

- Easy-to-use functions for creating scores, loadings, and biplots
- Tidyverse-compatible data structures
- Seamless integration with mixOmics objects (when installed)
- Flexible customization options
- Publication-quality plots

For more information, see the package documentation and the [GitHub repository](https://github.com/JannikVindeloev/ggmultivar).
